---
title: "Variable and value labels support in base R and other packages"
date: "`r Sys.Date()`"
output: 
    html_document
vignette: >
    %\VignetteIndexEntry{Variable and value labels support in base R and other packages}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{utf8}
---

'expss' package provides methods for labelled variables which add value labels support to base R functions and to some functions from other packages. Here we demonstrate labels support in base R - value labels automatically used as factors levels. Every function which internally converts variable to factor will utilize labels. 

First, apply value and variables labels to dataset:
```{r, message=FALSE, warning=FALSE}
library(expss)
data(mtcars)
mtcars = apply_labels(mtcars,
                      mpg = "Miles/(US) gallon",
                      cyl = "Number of cylinders",
                      disp = "Displacement (cu.in.)",
                      hp = "Gross horsepower",
                      drat = "Rear axle ratio",
                      wt = "Weight (1000 lbs)",
                      qsec = "1/4 mile time",
                      vs = "Engine",
                      vs = c("V-engine" = 0,
                             "Straight engine" = 1),
                      am = "Transmission",
                      am = c("Automatic" = 0,
                             "Manual"=1),
                      gear = "Number of forward gears",
                      carb = "Number of carburetors"
)

```

Base `table` and `boxplot` with value labels:
```{r, fig.height=6, fig.width=7}
with(mtcars, table(am, vs))

boxplot(mpg ~ am, data = mtcars)
```

There is a special function for variables labels support - `use_labels`. By now variables labels support available only for expression which will be evaluated inside data.frame.
```{r}
# table with dimension names
use_labels(mtcars, table(am, vs)) 

# linear regression
use_labels(mtcars, lm(mpg ~ wt + hp + qsec)) %>% summary
```

And, finally, `ggplot2` graphics with variables and value labels:
```{r, fig.height=6, fig.width=7}
library(ggplot2, warn.conflicts = FALSE)

use_labels(mtcars, {
    # '..data' is shortcut for all 'mtcars' data.frame inside expression 
    ggplot(..data) +
        geom_point(aes(y = mpg, x = wt, color = qsec)) +
        facet_grid(am ~ vs)
}) 
```
